---
title: "Introduction"
author: "Frederick De Baene"
format:
  html:
    toc: true
    number-sections: true
editor: visual
---

```{r echo=FALSE}
library(ggplot2)
```

# Introduction

This page serves as an introduction to different concepts and terms in
statistics.

# Population and Sample

The **population** is the complete set of elements in which we are interested
and on which we want to learn something. These elements can be individuals or
objects. We want to learn something about the characteristics of this
population, and we do this by studying the characteristics of the elements of
the population.

Because it is often not possible to study every element in the population, a
subset of the population is studies. This study is called a **sample**. The
sample must be representative in order to be able to conclude something about
the population from which it is drawn.

# Variables

We use **variables** to represent the characteristics of the elements. The
possible outcomes for a variable determine the type of variable we are dealing
with:

- quantitative: discrete or continuous
- qualitative: nominal or ordinal.

The possible outcomes for a discrete random variable are countable or finite, so
often the observed value for a discrete random variable is the result of
counting something. A continuous random variable has an infinite number of
possible outcomes, i.e., all the possible numbers within a particular range.

The possible outcomes for qualitative random variables are categories. If the
categories can be ordered, then we are talking about an ordinal random variable,
otherwise it is a nominal random variable.

Another definition of an experiment states that it captures the outcome of a
random experiment (random process). An example of a random experiment is throwing
a fair coin. The sample space is:

$$
S = \{ heads, tails \}
$$

Repeated executions of the same random experiment do not necessarily yield the
same result. If we execute the same experiment $k$ times, then the random
variable $Y_i$ captures the outcome of the $i$th execution of these $k$
executions of the same experiment.

# Distributions

A distribution shows how the outcomes or values of a variable are distributed
across the elements in a sample or population. However, there is a distinction
between **theoretical** and **empirical** distributions. A theoretical
distribution shows how the possible outcomes of a variable are distributed
across the elements in the population; they are based on mathematical models and
reflect what we expect to see. An empirical distribution shows how the observed
outcomes of a variable are distributed across the elements in a sample; they are
based on what we observe.

As an example, let us look at tossing a fair die. First, we must define the
sample space. The sample space is the complete set of possible outcomes:

$$
S = \{ 1, 2, 3, 4, 5, 6 \}
$$

We assume that each possible outcome is equally likely. The theoretical
distribution can be visualized as:

```{r echo = FALSE}
dat <- tibble::tibble(x = 1L:6L, p = rep(1/6, 6))
ggplot(dat, aes(x = x, y = p)) +
  geom_col() +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Theoretical distribution of a fair die toss") +
  labs(x = "Outcome (number of eyes)", y = "Probability") +
  theme(plot.title = element_text(hjust = 0.5))
```

Now, let us simulate an experiment where we toss a fair die 30 times and each
time note the number of eyes thrown.

```{r}
# Ensure reproducibility
set.seed(456)

# Toss a fair die 30 times and note the outcome of each toss
dat <- tibble::tibble(x = sample(x = 1L:6L, size = 30, replace = TRUE))
```

We can now use the observed values to visualize the empirical distribution:

```{r echo = FALSE}
ggplot(dat, aes(x = x)) +
  geom_bar() +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Empirical distribution of a fair die toss") +
  labs(x = "Outcome (number of eyes)", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))
```

We see that the theoretical and empirical distributions are not entirely the
same. The theoretical distribution shows us that each possible outcome is
equally likely. Therefore, when repeating the experiment 30 times, we expect
that 5 out of 30 times a 1 is thrown, 5 out of 30 times a 2 is thrown, etc. In
the actual experiment, a 1 and 3 were thrown 5 times each, as expected; a 2 and
4 were thrown 4 times each, 1 less than expected; and a 5 and 6 were thrown 6
times each, 1 more than expected.

# Measures

We can use measures to describe and understand distributions. Measures of
central tendency, such as the mean, median, and mode, tell us something about
the gravity of a distribution, about its typical value, while measures of
dispersion (or variability), such as range, standard deviation, and variance,
tell us something about how the values of a variable vary across observations.

## Measures of Central Tendency

The **mean** is calculated as follows:

$$
\overline{X} = \frac{1}{n} \sum_{i = 1}^{n} X_i
$$

It acts as a balance point for the data, and it is sensitive to outliers. The
**median** is the middle value of the ordered dataset. If the number of observations
is odd, the median is the value of the observation at the following position in an
ordered dataset:

$$
\frac{n + 1}{2}
$$

If the number of observations is even, the median is the mean of the middle two
observations at positions $n/2$ and $(n/2) + 1$ in the ordered dataset. The
median is not sensitivi to outliers. The **mode** is the most frequently
occurring value in the dataset.

# Case Study

Assume the population of interest comprises all students enrolled at a
particular university. In this case, the individual elements of the population
are persons (individuals). We are interested in different characteristics about
this population:

- the mean height
- the median number of pets
- the most frequently occurring hair color.

We do this by studying the characteristics for the individual students and use
variables to represent these characteristics. Let the random variable $X$ denote
the height of a student in centimeters, let the random variable $Z$ denote
number of pets a student has, and let the random variable $V$ denote the hair
color of a student. The random variables $X$, $Z$, and $V$ are a continuous,
discrete, and nominal variables, respectively.

Let the random variable $X$ denote the height of the students in centimeters;
this is a continuous random variable, it can take on any possible numeric value
in a particular range. Assume we know the height for each student in our
population.

```{r}
# Simulate the heights (cm) for each student in the population
set.seed(123)
n <- 60e3

dat <- tibble::tibble(X = rnorm(n = n, mean = 172, sd = 5))
```

We can now use the observed values to visualize the empirical distribution of
$X$.

```{r}
ggplot(dat, aes(x = X)) +
  geom_histogram(bins = 50) +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(title = "Empirical distribution of length") +
  labs(x = "Length (centimeters)", y = "Frequency") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Appendix A {.unnumbered}

The estimator for the population mean:

$$
\overline{X} = \frac{1}{n} \sum_{i = 1}^{n} X_i
$$

The mean of the sum of two random variables $X$ and $Y$ is the sum of their
means:

$$
\begin{align}
\overline{X + Y} 
  &= \frac{1}{n} \cdot \bigg[ \sum_{i = 1}^{n} \big[ X_i + Y_i \big] \bigg] \\
  &= \frac{1}{n} \cdot \bigg[ \sum_{i = 1}^{n} X_i + \sum_{i = 1}^{n} Y_i \bigg] \\
  &= \frac{1}{n} \cdot \sum_{i = 1}^{n} X_i + \frac{1}{n} \cdot \sum_{i = 1}^{n} Y_i \\
  &= \overline{X} + \overline{Y}
\end{align}
$$

The mean of a variable scaled by a constant $c$ is the mean of the original
variable scaled by that same constant $c$:

$$
\begin{align}
\overline{cX} 
  &= \frac{1}{n} \cdot \sum_{i = 1}^{n} c X_i \\
  &= c \cdot \frac{1}{n} \cdot \sum_{i = 1}^{n} X_i \\
  &= c \cdot \overline{X}
\end{align}
$$
