---
title: "Introduction"
author: "Frederick De Baene"
format: html
editor: visual
---

```{r echo=FALSE}
library(ggplot2)
```

# Introduction

This page serves as an introduction to different concepts and terms in
statistics.

# Population and Sample

The **population** is the complete set of elements in which we are interested
and on which we want to learn something. These elements can be individuals or
objects. We want to learn something about the characteristics of this
population, and we do this by studying the characteristics of the elements of
the population.

Because it is often not possible to study every element in the population, a
subset of the population is studies. This study is called a **sample**. The
sample must be representative in order to be able to conclude something about
the population from which it is drawn.

# Variables

We use **variables** to represent the characteristics of the elements. The
possible outcomes for a variable determine the type of variable we are dealing
with:

- quantitative: discrete or continuous
- qualitative: nominal or ordinal.

The possible outcomes for a discrete random variable are countable or finite, so
often the observed value for a discrete random variable is the result of counting
something. A continuous random variable has an infinite number of possible
outcomes, i.e., all the possible numbers within a particular range.

The possible outcomes for qualitative random variables are categories. If the
categories can be ordered, then we are talking about an ordinal random variable,
otherwise it is a nominal random variable.

The characteristics of a population are statistical measures, measures of center
and measures of dispersion. These measures describe the distribution of a
variable (characteristic in the population). A distribution shows how the values
of a variable are distributed across the elements in the population. A
**theoretical distribution** shows how the possible values for a variable should
be distributed across the elements in a population (or sample), while an
**empirical distribution** shows how the observed values for a variable are
distributed across the elements in the sample (or population).

As an example, let us look at tossing a fair die. First, we must define the
sample space. The sample space is the complete set of possible outcomes:

$$
S = \{ 1, 2, 3, 4, 5, 6 \}
$$

We assume that each possible outcome is equally likely to occur when the die is
tossed. Therefore, the theoretical distribution can be visualized as:

```{r}
dat <- tibble::tibble(x = 1L:6L, p = rep(1/6, 6))
ggplot(dat, aes(x = x, y = p)) +
    geom_col() +
    theme_classic() +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = "Theoretical distribution of a fair die toss") +
    labs(x = "Outcome (number of eyes)", y = "Probability") +
    theme(plot.title = element_text(hjust = 0.5))
```

Now, let us simulate an experiment where we toss a fair die 30 times and each
time note the number of eyes thrown.

```{r}
set.seed(456)
dat <- tibble::tibble(x = sample(x = 1L:6L, size = 30, replace = TRUE))

dat
```


## Measures of Center

There are different measures of center. The first one is the **mean**, and is
calculated as follows:

$$
\overline{X} = \frac{1}{n} \sum_{i = 1}^{n} X_i
$$

The median is the value for the variable for which at least half of the observed
values are less than or equal to. The median is obtained by ordering all
observed values from smallest to greatest and take the middle value. If the
number of observed values is even, we take the middle two values and calculate
their average.

# Case Study

Assume the population of interest comprises all students enrolled at a
particular university. In this case, the individual elements of the population
are persons (individuals). We are interested in different characteristics about
this population:

- the mean height
- the median number of pets
- the most frequently occurring hair color.

We do this by studying the characteristics for the individual students and use
variables to represent these characteristics. Let the random variable $X$ denote
the height of a student in centimeters, let the random variable $Z$ denote
number of pets a student has, and let the random variable $V$ denote the hair
color of a student. The random variables $X$, $Z$, and $V$ are a continuous,
discrete, and nominal variables, respectively.

Let the random variable $X$ denote the height of the students in centimeters;
this is a continuous random variable, it can take on any possible numeric value
in a particular range. Assume we know the height for each student in our
population.

```{r}
# Simulate the heights (cm) for each student in the population
set.seed(123)
n <- 60e3

dat <- tibble::tibble(X = rnorm(n = n, mean = 172, sd = 5))
```

We can now use the observed values to visualize the empirical distribution of
$X$.

```{r}
ggplot(dat, aes(x = X)) +
    geom_histogram(bins = 50) +
    theme_classic() +
    scale_y_continuous(expand = c(0, 0)) +
    labs(title = "Empirical distribution of length") +
    labs(x = "Length (centimeters)", y = "Frequency") +
    theme(plot.title = element_text(hjust = 0.5))
```

# Appendix A

The estimator for the population mean:

$$
\overline{X} = \frac{1}{n} \sum_{i = 1}^{n} X_i
$$

The mean of the sum of two random variables $X$ and $Y$ is the sum of their
means:

$$
\begin{align}
\overline{X + Y} 
    &= \frac{1}{n} \cdot \bigg[ \sum_{i = 1}^{n} \big[ X_i + Y_i \big] \bigg] \\
    &= \frac{1}{n} \cdot \bigg[ \sum_{i = 1}^{n} X_i + \sum_{i = 1}^{n} Y_i \bigg] \\
    &= \frac{1}{n} \cdot \sum_{i = 1}^{n} X_i + \frac{1}{n} \cdot \sum_{i = 1}^{n} Y_i \\
    &= \overline{X} + \overline{Y}
\end{align}
$$

