---
title: "Survival"
author: "Frederick De Baene"
format: pdf
---

```{r echo=FALSE}
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(survival))
suppressPackageStartupMessages(library(survminer))
suppressPackageStartupMessages(library(tidyverse))
```

# Survival

The survival function $S(t)$ calculates the probability that the event occurs
after time $t$:

$$
S(t) = \mathbb{P} \big[ T > t \big]
$$

Given a sample dataset, we can estimate the survival function $S(t)$ with the
following estimator:

$$
\tilde{S}(t) = \frac{\sum I[T > t]}{n}
$$

with $I[T > t]$ an indicator variable which is $1$ if the event occurs after
time $t$ (and $0$ otherwise), $n$ the sample size, and $\sum I[T > t]$ the number
of observations in the dataset for which the event occurs after time $t$.

In this sense, the estimator represents the fraction of patients that is event-free
at time $t$.

## No Censoring

```{r}
# Data from the Ohio State University Bone Marrow Transplant Unit
dat <- tibble(
  date = c(1, 2, 2, 2, 3, 5, 6, 7, 8, 16, 17, 34),
  cens = c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1)
)
```

```{r echo=FALSE}
kable(dat) 
```

For the dataset, we can estimate $S(t)$ using $\tilde{S}(t)$:

```{r}
# Use a naive non-parametric estimator to estimate the survival function
observation_months <- 0L:34L
S_tilde <- rep(NA, max(observation_months))

for (t in observation_months) {
  S_tilde[t + 1] <- cumprod(sum(dat$date > t) / nrow(dat))
}

km_estimates <- tibble(t = observation_months, S_tilde = S_tilde)
```

To estimate the true survival function $S(t)$, there is also the non-parametric
Kaplan-Meier (KM) estimator:

$$
\hat{S}(t) = \prod_{j; t_j < t} (1 - \frac{d_j}{r_j})
$$

with $r_j$ the number at risk at time $t_j$, $d_j$ the number of events that occur
at time $t_j$.

```{r}
# For each time interval (month), get the number at risk, the number of events,
# and the number of censored observations
r_j <- rep(NA, max(observation_months))
d_j <- rep(NA, max(observation_months))
c_j <- rep(NA, max(observation_months))

for (t in observation_months) {
  
  r_j[t + 1] <- sum(dat$date >= t)
  d_j[t + 1] <- sum(dat$date == t & dat$cens == 1)
  c_j[t + 1] <- sum(dat$date == t & dat$cens == 0)
  
}

km_estimates <- add_column(
  km_estimates,
  r_j = r_j,
  d_j = d_j,
  c_j = c_j,
  .before = "S_tilde"
)

# Calculate the hazard and estimate S(t)
km_estimates$hazard <- km_estimates$d_j / km_estimates$r_j
km_estimates$S_t <- cumprod(1 - km_estimates$hazard)
```

If there is no censoring, then $\tilde{S}(t)$ and $\hat{S}(t)$ are the same.

```{r echo=FALSE}
kable(km_estimates)
```

## With Censoring

If there is censoring, then the estimator $\tilde{S}(t)$ underestimates the
probability of survival. People who are censored at time $t$ do not die at
time $t$. However, $\tilde{S}(t)$ considers them as having had the event at
time $t$.

```{r}
# Data from the Ohio State University Bone Marrow Transplant Unit
dat <- tibble(
  date = c(1, 2, 2, 2, 3, 5, 6, 7, 8, 16, 17, 34),
  cens = c(1, 1, 1, 0, 1, 1, 1, 0, 1, 0, 1, 0)
)
```

The following patients were followed up for 34 months.

```{r echo=FALSE}
kable(dat)
```

```{r}
# Use a naive non-parametric estimator to estimate the survival function
observation_months <- 0L:34L
S_tilde <- rep(NA, max(observation_months))

for (t in observation_months) {
  S_tilde[t + 1] <- cumprod(sum(dat$date > t) / nrow(dat))
}

km_estimates <- tibble(t = observation_months, S_tilde = S_tilde)
```

```{r}
# For each time interval (month), get the number at risk, the number of events,
# and the number of censored observations
r_j <- rep(NA, max(observation_months))
d_j <- rep(NA, max(observation_months))
c_j <- rep(NA, max(observation_months))

for (t in observation_months) {
  
  r_j[t + 1] <- sum(dat$date >= t)
  d_j[t + 1] <- sum(dat$date == t & dat$cens == 1)
  c_j[t + 1] <- sum(dat$date == t & dat$cens == 0)
  
}

km_estimates <- add_column(
  km_estimates,
  r_j = r_j,
  d_j = d_j,
  c_j = c_j,
  .before = "S_tilde"
)

# Calculate the hazard and estimate S(t)
km_estimates$hazard <- km_estimates$d_j / km_estimates$r_j
km_estimates$S_t <- cumprod(1 - km_estimates$hazard)
```

# Hazard

The hazard is the instantaneous risk of the event occurring. Heuristically, we
can interpret this as a conditional probability of the event occurring at time
$t$ given that it has not yet occurred up to time $t$:

$$
\lambda_t = \mathbb{P} \big[ T = t | T \geq t \big] = \frac{d_t}{r_t}
$$

with $r_t$ the number at risk at time $t$ and $d_t$ the number of events that
occurred at time $t$.

```{r}
# Calculate the hazard
km_estimates$hazard <- km_estimates$d_j / km_estimates$r_j
```

We can now estimate the survival function with the non-parametric Kaplan-Meier
estimator.

```{r}
# Estimate S(t)
km_estimates$S_t <- cumprod(1 - km_estimates$hazard)
```

We can also estimate it using standard R packages.

```{r}
# Estimate S(t) using the non-parametric KM estimator
km_fit <- survfit(Surv(date, cens) ~ 1, data = dat)
```

```{r}
summary(km_fit)
```

We can plot the estimated survival curve.

```{r}
ggsurvplot(km_fit)
```

If we ignore censoring, then we underestimate the survival probability. Individuals
who are censored at time $t$ are now considered to have the event at time $t$.

```{r}
# Faulty estimation of overall survival
S_tilde_t <- rep(NA, max(observation_months))

for (t in observation_months) {
  S_tilde_t[t + 1] <- cumprod(sum(dat$date > t) / nrow(dat))
}

km_estimates$S_tilde_t <- S_tilde_t
```

```{r echo=FALSE}
kable(km_estimates)
```

# Standard Error

The Greenwood formula can be used to estimate the standard error of survival
probability estimates:

$$
\hat{\sigma^2} \big[ \hat{S}(t) \big] 
  = \big[ \hat{S}(t) \big] ^2 \cdot \sum_{t_i \leq t} \frac{d_i}{r_i (r_i - d_i)}
$$

We can calculate this manually.

$$
SE \big[ \hat{L}(t) \big] 
  = \frac
    { SE \big[ \hat{S}(t) \big] }
    { \hat{S}(t) \cdot | log \big( \hat{S}(t) \big) | }
$$

```{r}
# Calculate the estimated standard error using the Greenwood formula
fac_1 <- km_estimates$S_t^2
fac_2 <- cumsum(
  km_estimates$d_j
  / (km_estimates$r_j * (km_estimates$r_j - km_estimates$d_j))
)

km_estimates <- km_estimates |>
  add_column(est_var = fac_1 * fac_2) |>
  mutate(
    est_se = sqrt(est_var),
    conf_lower = S_t - 1.96*est_se,
    conf_upper = S_t + 1.96*est_se,
    adj_conf_lower = pmax(0, conf_lower),
    adj_conf_upper = pmin(1, conf_upper)
  )
```

```{r}
# Estimate S(t) using the non-parametric KM estimator
km_fit <- survfit(
  Surv(date, cens) ~ 1,
  data = dat,
  conf.type = "plain",
  conf.lower = "usual"
)
```

```{r}
summary(km_fit)
```

Using this approach, we can yield lower confidence limits smaller than 0 or upper
confidence limits greater than 1. We transform $\hat{S}(t)$:

$$
\hat{L}(t) = log(-log(S(t)))
$$

We estimate the standard error of $\hat{L}(t)$ as follows:

```{r}
L_t <- log(-log(km_estimates$S_t))
L_t_se <- km_estimates$est_se / (km_estimates$S_t * abs(log(km_estimates$S_t)))
L_t_lower <- L_t - 1.96*L_t_se
L_t_upper <- L_t + 1.96*L_t_se

S_t_lower <- exp(-exp(L_t_upper))
S_t_upper <- exp(-exp(L_t_lower))

S_t_lower
```

```{r}
# Estimate S(t) using the non-parametric KM estimator
km_fit <- survfit(
  Surv(date, cens) ~ 1,
  data = dat,
  conf.type = "log-log"
)
```

```{r}
summary(km_fit)
```

# Comparing Curves

```{r}
# Get a sample dataset
# Create a data frame
dat <- data.frame(
  time = c(12, 18, 24, 30, 36, 6, 15, 20, 28, 40, 8, 10, 14, 16, 22, 5, 12, 18, 20, 25),
  status = c(1, 1, 0, 1, 0, 1, 1, 0, 1, 0, 1, 1, 1, 1, 0, 1, 1, 0, 1, 0),
  group = factor(rep(c("Treatment", "Control"), each = 10))
)
```

```{r}
km_fit <- survfit(Surv(time, status) ~ group, data = dat)
```

```{r}
summary(km_fit)
```

```{r}
dat_split <- split(dat, dat$group)
```

Adding a covariate to `survfit()` just splits the data in two to estimate the
survival curves for both populations using the respective samples.

```{r}
summary(survfit(Surv(time, status) ~ 1, data = dat_split[[1]]))
```

```{r}
ggsurvplot(km_fit)
```

```{r}
# Compare survival in both groups
surv_diff <- survdiff(Surv(time, status) ~ group, data = dat)
```

```{r}
surv_diff$var
```

```{r}
# For each time interval (month), get the number at risk, the number of events,
# and the number of censored observations
observation_months <- seq(0L, max(dat$time))

r_j <- rep(NA, max(observation_months))
d_j <- rep(NA, max(observation_months))
c_j <- rep(NA, max(observation_months))

r_j_trt <- rep(NA, max(observation_months))
d_j_trt <- rep(NA, max(observation_months))
c_j_trt <- rep(NA, max(observation_months))

r_j_ctl <- rep(NA, max(observation_months))
d_j_ctl <- rep(NA, max(observation_months))
c_j_ctl <- rep(NA, max(observation_months))

for (t in observation_months) {
  
  r_j[t + 1] <- sum(dat$time >= t)
  d_j[t + 1] <- sum(dat$time == t & dat$status == 1)
  c_j[t + 1] <- sum(dat$time == t & dat$status == 0)
  
  r_j_trt[t + 1] <- sum(dat$time >= t & dat$group == "Treatment")
  d_j_trt[t + 1] <- sum(dat$time == t & dat$group == "Treatment" & dat$status == 1)
  c_j_trt[t + 1] <- sum(dat$time == t & dat$group == "Treatment" & dat$status == 0)
  
  r_j_ctl[t + 1] <- sum(dat$time >= t & dat$group == "Control")
  d_j_ctl[t + 1] <- sum(dat$time == t & dat$group == "Control" & dat$status == 1)
  c_j_ctl[t + 1] <- sum(dat$time == t & dat$group == "Control" & dat$status == 0)
  
}

km_estimates <- tibble(
  time = observation_months,
  r_j = r_j,
  d_j = d_j,
  c_j = c_j,
  r_j_trt = r_j_trt,
  d_j_trt = d_j_trt,
  c_j_trt = c_j_trt,
  r_j_ctl = r_j_ctl,
  d_j_ctl = d_j_ctl,
  c_j_ctl = c_j_ctl
)

# Calculate the expected number of events for each group at each time point
km_estimates <- km_estimates |>
  mutate(
    d_j_trt_exp = r_j_trt * (d_j / r_j),
    v_j_trt = (r_j_trt * r_j_ctl * d_j * (r_j - d_j)) / (r_j^2 * (r_j - 1)),
    d_j_ctl_exp = r_j_ctl * (d_j / r_j),
    v_j_ctl = (r_j_trt * r_j_ctl * d_j * (r_j - d_j)) / (r_j^2 * (r_j - 1))
  ) |>
  filter(d_j > 0)
```

```{r}
sum(km_estimates$d_j_ctl)
sum(km_estimates$d_j_ctl_exp)
(sum(km_estimates$d_j_ctl) - sum(km_estimates$d_j_ctl_exp))^2 / sum(km_estimates$v_j_ctl, na.rm = TRUE)
```

```{r}
sum(km_estimates$d_j_trt)
sum(km_estimates$d_j_trt_exp)
(sum(km_estimates$d_j_trt - km_estimates$d_j_trt_exp))^2 / sum(km_estimates$v_j_trt, na.rm = TRUE)
```
