---
title: "Survival: Time-Varying Covariates"
author: "Frederick De Baene"
format:
  html:
    toc: true
    number-sections: true
---

```{r echo=FALSE}
suppressPackageStartupMessages(library(survsim))
suppressPackageStartupMessages(library(survival))
```

The below code generates a dataset with two time-fixed covariates: age and group. Furthermore,
there is also the event indicator **status** and the outcome **stop** (the time variable).

```{r}
# Initialize settings for simulation
set.seed(123)
N <- 100

# Simulate data
dat_tf <- simple.surv.sim(
  n = N,
  foltime = 500,
  dist.ev = c('llogistic'),
  anc.ev = c(0.68),
  beta0.ev = c(5.8),
  anc.cens = 1.2,
  beta0.cens = 7.4,
  z = list(c("unif", 0.8, 1.2)),
  beta = list(c(-0.4), c(0)),
  x = list(c("bern", 0.5), c("normal", 70, 13))
)
names(dat_tf)[c(1,6,7)] <- c("id", "group", "age")
```

The following also generates a dataset for 100 patients with the time-varying
covariate **crp** (C-reactive protein). The dataset follows a counting process
data structure. Such data structures allow for:

* time-varying covariates
* multiple events per subject
* left-truncated data
* interval-censored data.

```{r}
# Initialize settings for simulation
set.seed(123)
N <- 100
nft <- sample(x = 1:10, size = N, replace = TRUE) # number of follow-up time points
crp <- round(abs(rnorm(n = sum(nft) + N, mean = 100, sd = 40)), 1)
time <- NA
id <- NA
i <- 0

# Simulate data
for(n in nft) {
  i <- i+1
  time.n <- sample(1:500, n)
  time.n <- c(0, sort(time.n))
  time <- c(time, time.n)
  id.n <- rep(i,n+1)
  id <- c(id, id.n)
}
dat_tv <- cbind(data.frame(id, time)[-1,], crp)
```

We merge the datasets to obtain a start-stop dataset.

```{r}
dat <- tmerge(dat_tf, dat_tf, id = id, endpt = event(stop, status))
dat <- tmerge(dat, dat_tv, id = id, crp = tdc(time, crp))
```

Now, let's fit a Cox model with time-varying covariates.

```{r}
cox_fit <- coxph(
  Surv(tstart, tstop, endpt) ~ group + age + crp + cluster(id),
  data = dat
)
```

```{r echo=FALSE}
summary(cox_fit)
```

# Test PH Assumption

We can test the PH assumption as follows.

```{r}
# Fit the Cox model
cox_fit_lung <- coxph(Surv(time, status) ~ age + ph.karno + sex, data = lung)
zph <- cox.zph(cox_fit_lung)
```

```{r echo=FALSE}
zph
```

There is a significant deviation from the PH assumption for the variable
**ph.karno**, which shows $p = 0.0046$.

```{r}
plot(zph[2],lwd=2)
abline(0,0, col=1,lty=3,lwd=2)
abline(h= cox_fit_lung$coef[2], col=3, lwd=2, lty=2)
legend(
  "bottomright",
  legend = c(
    "Reference line for null effect",
    "Average hazard over time",
    "Time-varying hazard"
  ),
  lty = c(3,2,1),
  col = c(1,3,1),
  lwd = 2
)
```

# Time-Fixed Covariates

We can also assume that the effect of time-fixed covariates (or baseline covariates)
changes in time. We can split the observation time into several time intervals and assess
the effect of the time-fixed covariate during each of these time intervals.

```{r}
# Split the dataset into multiple time intervals
lung.split <- survSplit(
  Surv(time, status) ~ .,
  data = lung,
  cut = c(180, 350),
  episode = "tgroup",
  id = "id"
)
```

```{r}
# Fit a Cox model to the data
cox_fit_lung_split <- coxph(
  Surv(tstart, time, status) ~ age + ph.karno:strata(tgroup) + sex,
  data = lung.split
)
```

```{r echo=FALSE}
summary(cox_fit_lung_split)
```

```{r}
cox.zph(cox_fit_lung_split)
```

We can also model a covariate as a function of the covariate in function
of time.

```{r}
coxph(
  Surv(time, status) ~ age + ph.karno + tt(ph.karno) + sex,
  data = lung,
  tt = function(x, t, ...) x * log(t + 20)
)
```

The time-varying effect of **ph.karno** is:

$$
\beta_{ph.karno}(t) = -0.097932 + 0.015236 \cdot log(t + 20)
$$
