---
title: "Variance"
author: "Frederick De Baene"
format:
  html:
    toc: true
---

```{r echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
```

# Variance Estimation

Assume we are studying two populations. The random variable $X$ is distributed
normally in both populations. In population 1:

$$
X \sim N(\mu_1, \sigma^2)
$$

with $\mu_1 = 15$. In population 2:

$$
X \sim N(\mu_2, \sigma^2)
$$

with $\mu_2 = 25$. Note that in both populations the variance $\sigma^2$ of X
is the same, i.e., $\sigma^2 = 25$. Two populations with the same variance is denoted
by homoscedasticity. We can use this to our advantage to obtain more precise estimates
of $\sigma^2$.

For this simulation, we repeatedly sample 15 observations from each population.
Let $n_1$ ($n_2$) denote the sample size of the sample randomly drawn from
population 1 (2). For each sample, we use the variance estimator $\hat{\sigma}(X)$
to estimate the population variance for each population separately:

$$
\hat{\sigma}^2 = \frac{\sum_{i = 1}^{n} (X_i - \overline{X})^2}{n - 1}
$$

Then we also use the estimator for the pooled variance:

$$
\hat{\sigma}_{pooled}^2 = \frac
  {(n_1 - 1) \cdot s_1^2 + (n_2 - 1) \cdot s_2^2}
  {n_1 + n_2 - 2}
$$

with $s_1^2$ and $s_2^2$ the sample standard deviations for the samples from
population 1 and 2, respectively.

```{r}
# Define settings for simulation
set.seed(4294)

n_1 <- n_2 <- 15
N <- 5000L

mu_1 <- 15
mu_2 <- 25
var_1 <- 25
var_2 <- 25

est_var_1 <- rep(NA, N)
est_var_2 <- rep(NA, N)
est_pool_var <- rep(NA, N)

for (i in 1L:N) {
  
  # Randomly draw samples from populations 1 and 2
  X_1 <- rnorm(n = n_1, mean = mu_1, sd = sqrt(var_1))
  X_2 <- rnorm(n = n_2, mean = mu_2, sd = sqrt(var_2))
  
  # Estimate the population variance in both populations separately
  est_var_1[i] <- var(X_1)
  est_var_2[i] <- var(X_2)
  
  # Estimate pooled variance
  est_pool_var[i] <- ((n_1 - 1)*var(X_1) + (n_2 - 1)*var(X_2)) / (n_1 + n_2 - 2)
  
}

sim_res <- tibble(
  idx = 1L:N,
  est_var_1 = est_var_1,
  est_var_2 = est_var_2,
  est_pool_var = est_pool_var
)

sim_res <- pivot_longer(sim_res, cols = c(est_var_1, est_var_2, est_pool_var))
```

We can see that the precision of the pooled variance estimator is better than
the precision of the estimators for the variances for each population.

```{r echo=FALSE}
ggplot(sim_res, aes(x = value, color = name, fill = name)) +
  geom_density(alpha = 0.3) +
  geom_vline(xintercept = var_1, color = "red") +
  theme_classic() +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "Distribution of variance estimates",
    x = "Variance estimate", y = NULL
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```
